---
title: "Hands-on Exercise 9A"
author: "ZHANG Xiaoyu"
date: "Jun 18, 2026"
date-modified: "Jun 19,2026"
format: html
code-link: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **Choropleth Mapping with R**

Choropleth maps are widely used to visualize the spatial distribution of demographic, social, and economic variables. By applying graduated colors to geographic units, they allow users to easily identify spatial patterns and regional variations. In this exercise, the **tmap** package in R is used to construct choropleth maps and explore best practices for effective thematic mapping.

## **1. Getting Started**

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## **2. Importing Data into R**

This exercise uses two datasets.

The first is a shapefile containing the boundaries of Singapore’s planning subzones, which provides the geographic information needed for mapping.

The second is a population dataset in CSV format that contains demographic information such as age group, gender, and dwelling type. While the population data does not contain location coordinates, it can be joined to the shapefile using the Planning Area (PA) and Subzone (SZ) fields.

After combining the two datasets, choropleth maps can be created to show how population characteristics vary across different subzones in Singapore.

```{r}
mpsz <- st_read(dsn = "data/MP14_SUBZONE_WEB_PL", 
                layer = "MP14_SUBZONE_WEB_PL")
```

```{r}
mpsz
```

```{r}
popdata <- read_csv("data/respopagesextod2011to2020.csv")
```

## **3. Data Preparation**

Before mapping, the population data was cleaned and transformed to create a 2020 summary table. New variables were derived to represent the young population (0–24 years), economy-active population (25–64 years), aged population (65+ years), total population, and dependency ratio. Data wrangling was performed using functions from the **dplyr** and **tidyr** packages, including filtering, grouping, reshaping, and variable creation.

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

Before performing the georelational join, the values in the **PA** and **SZ** fields were converted to uppercase to ensure consistency with the shapefile, where **PLN_AREA_N** and **SUBZONE_N** are stored in uppercase. This preprocessing step helped avoid mismatches when joining the two datasets.

```{r}
popdata2020 <- popdata2020 %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, we use `left_join()` from dplyr to combine the geographical data with the attribute table. We match them using the planning subzone names (for example, `SUBZONE_N` in one dataset and `SZ` in the other) as the common key.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

## **4. Choropleth Mapping Geospatial Data Using *tmap***

### **4.1 Plotting a choropleth map quickly by using *qtm()***

The easiest way to draw a choropleth map in **tmap** is to use `qtm()`. It is short and simple to write, and it often gives a good default map without needing many extra settings. The code below creates a standard choropleth map, as shown in the example.

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

### **4.2 Creating a choropleth map by using *tmap*’s elements**

Although `qtm()` is very convenient for quickly creating a choropleth map, it has a limitation: it is difficult to fine-tune the appearance of individual map layers.

Therefore, if you want to produce a more polished, publication-quality choropleth map like the one shown below, you should use the full set of drawing functions provided by **tmap** instead of `qtm()`.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

The basic structure of **tmap** starts with `tm_shape()`, followed by layer functions like `tm_polygons()`. Here, `tm_shape()` sets the data (`mpsz_pop2020`), and `tm_polygons()` draws the subzone polygons.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

To draw a choropleth map showing the geographical distribution of a selected variable by planning subzone, we just need to assign the target variable such as *Dependency* to *tm_polygons()*.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

In fact, `tm_polygons()` is a combination of `tm_fill()` and `tm_borders()`.

`tm_fill()` colors the polygons using a default color scheme

`tm_borders()` adds the polygon outlines

The code below shows how to draw a choropleth map using only `tm_fill()`.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

To add the outlines of the subzones, we use `tm_borders()`, as shown in the code below.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY") +
  tm_borders(lwd = 0.01,  
             fill_alpha = 0.1)
```

### **4.3 Data classification methods of tmap**

Most choropleth maps use **data classification** to group many values into a few ranges (classes). In **tmap**, there are 10 classification methods, including: fixed, sd, equal, pretty (default), quantile, kmeans, hclust, bclust, fisher, and jenks.

The code below uses the **quantile** method with 5 classes.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

In the code chunk below, *equal* data classification method is used.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "equal",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

For built-in styles, class breaks are automatically calculated by **tmap**.

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

Now, we will plot the choropleth map by using the code chunk below.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(fill_alpha = 0.5)
```

### **4.4 Colour Scheme**

**tmap** supports color ramps that can either be custom-defined or taken from preset palettes in the **RColorBrewer** package.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

Notice that the choropleth map is shaded in green. To reverse the colour shading, add a “-” prefix.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

### **4.5 Map Layouts**

A map layout is how all map elements are combined into one complete map.

These elements include the data layer, title, scale bar, compass, margins, and aspect ratio.

The color settings and classification methods discussed earlier control how the map is styled, especially the colors and data breakpoints.

#### Map Legend

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5,
        values = "brewer.greens"),
      fill.legend = tm_legend(
        title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Dependency Ratio by planning subzone \n(Jenks classification)")
```

#### Map style

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

#### 5.3 Cartographic Furniture

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

### **4.6 Drawing Small Multiple Choropleth Maps**

Small multiple maps (or facet maps) are a set of maps shown side by side or stacked. They help compare how spatial patterns change across variables like time.

In **tmap**, you can create them in three ways:

-   Use multiple values in aesthetic arguments

-   Use `tm_facets()` to group data

-   Combine separate maps using `tmap_arrange()`

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal", 
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

In this example, small multiple choropleth maps are created by assigning multiple values to at least one of the aesthetic arguments

```{r}
tm_shape(mpsz_pop2020)+ 
  tm_polygons(c("DEPENDENCY","AGED"),
          style = c("equal", "quantile"), 
          palette = list("Blues","Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

In this example, multiple small choropleth maps are created by using `tm_facets()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) + 
  tm_facets(by="REGION_N", 
            free.coords=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

In this example, multiple small choropleth maps are created by creating multiple stand-alone maps with `tmap_arrange()`.

```{r}
youngmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("YOUNG", 
              style = "quantile", 
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("AGED", 
              style = "quantile", 
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

### **4.7 Mappping Spatial Object Meeting a Selection Criterion**

Instead of creating small multiple choropleth maps, you can use the selection function to map spatial objects that meet a specific criterion.

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45, 
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```
