---
title: "Hands-on Exercise 9B"
author: "ZHANG Xiaoyu"
date: "Jun 18, 2026"
date-modified: "Jun 19,2026"
format: html
code-link: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **Visualising Geospatial Point Data**

Proportional symbol maps (or graduated symbol maps) use the size of symbols to show differences in values, such as the number of people.

We can create both classed and unclassed maps. Classed maps are called graduated symbols, while unclassed maps are proportional symbols, where symbol size is directly proportional to the values.

In this exercise, you will use the **tmap** package to create a proportional symbol map showing the number of wins at Singapore Pools outlets.

## **1. Getting Started**

Before we get started, we need to ensure that **tmap** package of R and other related R packages have been installed and loaded into R.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## **2. Data Import and Preparation**

The code chunk below uses *read_csv()* function of **readr** package to import *SGPools_svy21.csv* into R as a tibble data frame called *sgpools*.

```{r}
sgpools <- read_csv("data/SGPools_svy21.csv")
```

The code chunk below shows list() is used to do the job.

```{r}
list(sgpools) 
```

## **3. Creating a sf data frame from an aspatial data frame**

The code chunk below converts sgpools data frame into a simple feature data frame by using *st_as_sf()* of **sf** packages

```{r}
sgpools_sf <- st_as_sf(sgpools, 
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

We can display the basic information of the newly created sgpools_sf by using the code chunk below.

```{r}
list(sgpools_sf)
```

## **4. interactive point symbol map**

The code chunks below are used to create an interactive point symbol map.

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "red",
           size = 1,
           col = "black",
           lwd = 1)
```

To draw a proportional symbol map, we need to assign a numerical variable to the size visual attribute. The code chunks below show that the variable *Gp1Gp2Winnings* is assigned to size visual attribute.

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "red",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

The proportional symbol map can be further improved by using the colour visual attribute. In the code chunks below, *OUTLET_TYPE* variable is used as the colour attribute variable.

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", 
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

An impressive and little-know feature of **tmap**’s view mode is that it also works with faceted plots. The argument *sync* in *tm_facets()* can be used in this case to produce multiple maps with synchronised zoom and pan settings.

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", 
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1) + 
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```
