---
title: "Hands-on Exercise 4B"
author: "ZHANG Xiaoyu"
date: "May 07, 2026"
date-modified: "May 07,2026"
format: html
code-link: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **Visual Statistical Analysis**

Statistical inference is a very useful statistical method. However, it can sometimes be difficult and abstract. In this section, we will use data visualisation to make statistical inference easier to understand and more intuitive.

**ggstatsplot** is an extension of **ggplot2**. It creates plots that include statistical test results.

## **1. Installing and launching R packages**

```{r}
pacman::p_load(ggstatsplot, tidyverse)
```

## **2. Importing data**

In this chapter we still use the *Exam_data.csv*

```{r}
exam_data <- read_csv("data/Exam_data.csv")

```

## **3. One-sample test: *gghistostats()* method**

A one-sample test is used to compare the mean of one sample with a known or expected value.

It helps us check whether the sample result is significantly different from that value.

Next, we use [*gghistostats()*](https://indrajeetpatil.github.io/ggstatsplot/reference/gghistostats.html) to build an visual of one-sample test on English scores.

::: panel-tabset
### The plot

```{r}
set.seed(1234)
exam_data$ENGLISH <- as.numeric(exam_data$ENGLISH)

gghistostats(
  data = exam_data,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores"
)
```

### The code chunk

```{r}
#| eval: false
set.seed(1234)
exam_data$ENGLISH <- as.numeric(exam_data$ENGLISH)

gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores"
)
```
:::

## **4. Unpacking the Bayes Factor**

A Bayes Factor compares two hypotheses. It shows which hypothesis is better supported by the data.

When comparing the alternative hypothesis (**H1**) and the null hypothesis (**H0**), the Bayes Factor is written as **B10**.

A larger Bayes Factor means stronger evidence for **H1**. A smaller value means stronger evidence for **H0**.

The Schwarz criterion is a simple way to estimate the Bayes Factor.Bayes Factor values can then be used to measure the strength of evidence.

## **5. Two-sample mean test: *ggbetweenstats()***

```{r}
exam_data$MATHS <- as.numeric(exam_data$MATHS)
ggbetweenstats(
  data = exam_data,
  x = GENDER, 
  y = MATHS,
  type = "np",
  messages = FALSE
)
```

## **6. Oneway ANOVA Test: *ggbetweenstats()* method**

```{r}
ggbetweenstats(
  data = exam_data,
  x = RACE, 
  y = ENGLISH,
  type = "p",
  mean.ci = TRUE, 
  pairwise.comparisons = TRUE, 
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```

## **7. Significant Test of Correlation: *ggscatterstats()***

```{r}
ggscatterstats(
  data = exam_data,
  x = MATHS,
  y = ENGLISH,
  marginal = FALSE,
  )
```

## **8. Significant Test of Association (Depedence) : *ggbarstats()* methods**

```{r}
exam1 <- exam_data %>% 
  mutate(MATHS_bins = 
           cut(MATHS, 
               breaks = c(0,60,75,85,100))
)
```

```{r}
ggbarstats(exam1, 
           x = MATHS_bins, 
           y = GENDER)
```
