---
title: "Hands-on Exercise 1"
author: "ZHANG Xiaoyu"
date: "April 17, 2026"
date-modified: "April 18,2026"
code-link: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Getting Started

## Install and launching R packages

The code chunk below uses

\[p_load()\](https://www.rdocumentation.org/packages/pacman/versions/0.5.1/topics/p_load) of pacman package to check if tidyverse packages are installed in this computer. If they are,then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## Plotting a simple bar chart

```{r}
ggplot(data = exam_data, 
       aes(x = fct_infreq(RACE))) + 
  geom_bar()

```

# **R Graphics VS ggplot**

## **R Graphics**

```{r}
exam_data$MATHS <- as.numeric(exam_data$MATHS)
hist(exam_data$MATHS)
```

## **ggplot**

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10, 
                 boundary = 100,
                 color="black", 
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```

# **Essential Grammatical Elements in ggplot2**

## Call the `ggplot()` function

```{r}
ggplot(data=exam_data)
```

## Adds the aesthetic element

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS))
```

## Plot a bar chart using `geom_bar()`

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

## Plot a dot plot using [`geom_dotplot()`](https://ggplot2.tidyverse.org/reference/geom_dotplot.html)

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL) 
```

## Create a simple histogram using [*geom_histogram()*](https://ggplot2.tidyverse.org/reference/geom_histogram.html)

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram() 
```

## **Modifying a geometric object by changing `geom()`**

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue") 
```

## **Modifying a geometric object by changing *aes()***

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```

## Plot the distribution of Maths scores using [`geom-density()`](https://ggplot2.tidyverse.org/reference/geom_density.html)

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()  
```

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```

## Plot boxplots using [`geom_boxplot()`](https://ggplot2.tidyverse.org/reference/geom_boxplot.html)

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot() 
```

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

## Plot the distribution of Maths score by gender using [`geom_violin`](https://ggplot2.tidyverse.org/reference/geom_violin.html)

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

## Plot scatterplot using `geom_point()`

```{r}
exam_data$ENGLISH <- as.numeric(exam_data$ENGLISH)

ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  scale_y_continuous(breaks = seq(0, 100, by = 20))
```

## Plot the data points on the boxplots by using both `geom_boxplot()` and `geom_point()`

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```

## Add mean values by using [`stat_summary()`](https://ggplot2.tidyverse.org/reference/stat_summary.html)

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)               
```

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun="mean",           
             colour="red",          
             size=4)          
```

## **Adding a best fit curve on a scatterplot**

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```

## **Working with `facet_wrap()`**

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

## Plot a trellis plot using `facet_grid()`

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
  facet_grid(~ CLASS) +
  scale_x_continuous(breaks = seq(0, 100, by = 40))
```

## Flip the horizontal bar chart into vertical bar chart by using `coord_flip()`

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

## Fix both the y-axis and x-axis range from 0-100

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

## Plot a horizontal bar chart using `theme_gray()`

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```
