---
title: "Hands-on Exercise 5B"
author: "ZHANG Xiaoyu"
date: "May 12, 2026"
date-modified: "May 12,2026"
format: html
code-link: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **Visual Correlation Analysis**

**Correlation coefficients** (ranging from -1.0 to 1.0) measure the strength and direction of the relationship between two variables. When dealing with multiple variables, we organize these values into a **correlation matrix**.

## **1. Installing and Launching R Packages**

Here are the R packages we will use:

-   **`tidyverse`**: For data cleaning and preparation.

-   **`corrplot`**: For creating visual, color-coded correlation maps (corrgrams).

-   **`ggpubr`**: For generating professional, publication-ready charts.

-   **`plotly`**: For building interactive versions of our correlation plots.

```{r}
pacman::p_load(corrplot, ggstatsplot, tidyverse)
```

## **2. Importing and Preparing The Data Set**

In this hands-on exercise, the [Wine Quality Data Set](https://archive.ics.uci.edu/ml/datasets/wine+quality) of UCI Machine Learning Repository will be used.

```{r}
wine <- read_csv("data/wine_quality.csv")
```

## **3. Building Correlation Matrix: *pairs()* method**

In a scatterplot matrix, we compare every variable in the dataset against every other variable. Each small box in the grid is an individual **scatterplot** showing the correlation between two specific wine characteristics (like alcohol vs. pH).

```{r}
pairs(wine[,1:11])
```

But this plot seems messy,because we are trying to compare 11 or 12 variables at once, which creates a huge grid with very tiny plots.

Then we try to display the relationships among variables: fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, free sulfur dioxide, total sulfur dioxide, density, pH, sulphates and alcohol.

```{r}
pairs(wine[,2:12])
```

### **Drawing the lower corner**

To let the plot seems more clear,we can try to show either the upper half or lower half of the correlation matrix instead of both. This is because a correlation matrix is symmetric.

```{r}
pairs(wine[,2:12], upper.panel = NULL)
```

Same as the upper side,we can only show the lower side:

```{r}
pairs(wine[,2:12], lower.panel = NULL)
```

### **Including with correlation coefficients**

Though we can not see the correlation coefficient in the above plot,that not convenient for us to do analysis,so we try to show the correlation coefficient of each pair of variables.

```{r}
panel.cor <- function(x, y, digits=2, prefix="", cex.cor, ...) {
usr <- par("usr")
on.exit(par(usr))
par(usr = c(0, 1, 0, 1))
r <- abs(cor(x, y, use="complete.obs"))
txt <- format(c(r, 0.123456789), digits=digits)[1]
txt <- paste(prefix, txt, sep="")
if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
text(0.5, 0.5, txt, cex = cex.cor * (1 + r) / 2)
}

pairs(wine[,2:12], 
      upper.panel = panel.cor)
```

## **4. Visualising Correlation Matrix: ggcormat()**

One of the major limitation of the correlation matrix is that the scatter plots appear very cluttered when the number of observations is relatively large (i.e. more than 500 observations).

Next we will use [*ggcorrmat()*](https://indrajeetpatil.github.io/ggstatsplot/reference/ggcorrmat.html) of [**ggstatsplot**](https://indrajeetpatil.github.io/ggstatsplot/index.html) package to conduct **Corrgram** data visualisation technique.

### **The basic plot**

```{r}
#| fig.width: 15
#| fig.height: 10
ggstatsplot::ggcorrmat(
  data = wine, 
  cor.vars = 1:11)
```

```{r}
#| fig.width: 15
#| fig.height: 10
ggstatsplot::ggcorrmat(
  data = wine, 
  cor.vars = 1:11,
  ggcorrplot.args = list(outline.color = "black", 
                         hc.order = TRUE,
                         tl.cex = 10),
  title    = "Correlogram for wine dataset",
  subtitle = "Four pairs are no significant at p < 0.05"
)
```

```{r}
ggplot.component = list(
    theme(text=element_text(size=5),
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(size = 8)))
```

## **5. Building multiple plots**

Since ggstasplot is an extension of ggplot2, it also supports faceting. However the feature is not available in *ggcorrmat()* but in the [*grouped_ggcorrmat()*](https://indrajeetpatil.github.io/ggstatsplot/reference/grouped_ggcorrmat.html) of **ggstatsplot**.

```{r}
#| fig.width: 15
#| fig.height: 10
grouped_ggcorrmat(
  data = wine,
  cor.vars = 1:11,
  grouping.var = type,
  type = "robust",
  p.adjust.method = "holm",
  plotgrid.args = list(ncol = 2),
  ggcorrplot.args = list(outline.color = "black", 
                         hc.order = TRUE,
                         tl.cex = 10),
  annotation.args = list(
    tag_levels = "a",
    title = "Correlogram for wine dataset",
    subtitle = "The measures are: alcohol, sulphates, fixed acidity, citric acid, chlorides, residual sugar, density, free sulfur dioxide and volatile acidity",
    caption = "Dataset: UCI Machine Learning Repository"
  )
)
```

## **6. Visualising Correlation Matrix using corrplot Package**

This exercise focuses on using the **`corrplot`** package.

### **Getting started with corrplot**

```{r}
wine.cor <- cor(wine[, 1:11])
```

```{r}
corrplot(wine.cor)
```

### **Working with visual geometrics**

The **`corrplot`** package offers seven visual methods to display correlation values: **circle**, **square**, **ellipse**, **number**, **shade**, **color**, and **pie**. While **circle** is the default setting, we can easily switch to a different style by using the `method` argument in our code.

```{r}
corrplot(wine.cor, 
         method = "ellipse") 
```

The current layout looks messy, we can reorganize the content for a cleaner presentation.

```{r}
corrplot(wine.cor, 
         method = "ellipse", 
         type="lower")
```

The default layout of the corrgram can be further customised.

```{r}
corrplot(wine.cor, 
         method = "ellipse", 
         type="lower",
         diag = FALSE,
         tl.col = "black")
```

### **Working with mixed layout**

With **corrplot** package, it is possible to design corrgram with mixed visual matrix of one half and numerical matrix on the other half.

```{r}
#| fig.width: 15
#| fig.height: 10
corrplot.mixed(wine.cor, 
               lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

### **Combining corrgram with the significant test**

With corrplot package, we can use the *cor.mtest()* to compute the p-values and confidence interval for each pair of variables.

```{r}
wine.sig = cor.mtest(wine.cor, conf.level= .95)
```

We can then use the *p.mat* argument of *corrplot* function as shown in the code chunk below.

```{r}
#| fig.width: 15
#| fig.height: 10
corrplot(wine.cor,
         method = "number",
         type = "lower",
         diag = FALSE,
         tl.col = "black",
         tl.srt = 45,
         p.mat = wine.sig$p,
         sig.level = .05)
```

### **Reorder a corrgram**

Reordering the matrix is key to finding hidden patterns in your data. While `corrplot` defaults to the original order, we can change this using the **`order`** argument.

Available sorting methods include:

-   **"AOE"**: Sorts by the angular order of eigenvectors.

-   **"FPC"**: Sorts by the first principal component.

-   **"hclust"**: Uses hierarchical clustering (you can also choose specific methods like "ward" or "average").

-   **"alphabet"**: Sorts variables alphabetically.

For even more options, we can use the **`seriation`** package.

```{r}
#| fig.width: 15
#| fig.height: 10
corrplot.mixed(wine.cor, 
               lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               order="AOE",
               tl.col = "black")
```

### **Reordering a correlation matrix using hclust**

```{r}
#| fig.width: 15
#| fig.height: 10
corrplot(wine.cor, 
         method = "ellipse", 
         tl.pos = "lt",
         tl.col = "black",
         order="hclust",
         hclust.method = "ward.D",
         addrect = 3)
```
