---
title: "Hands-on Exercise 5D"
author: "ZHANG Xiaoyu"
date: "May 13, 2026"
date-modified: "May 14,2026"
format: html
code-link: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **Visual Multivariate Analysis with Parallel Coordinates Plot**

Parallel coordinates plots are used to visualize and analyze multivariate numerical data. They help compare multiple variables at once and reveal relationships between them, such as factors in a Happiness Index.

## **1. Installing and Launching R Packages**

We will use the **GGally**, **parallelPlot** and **tidyverse** package in this chapter.

```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```

## **2. Data Preparation**

In this exercise, we still use the World Happiness 2018 data set.

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

## **3. Plotting Static Parallel Coordinates Plot**

In this section we will create static parallel coordinates plots using `ggparcoord()` from the `GGally` package.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12))
```

### **Plotting a parallel coordinates with boxplot**

```{r}
#| fig.width: 15
#| fig.height: 10
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")
```

### **Parallel coordinates with facet**

```{r}
#| fig.width: 15
#| fig.height: 10
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)
```

### **Rotating x-axis text label**

```{r}
#| fig.width: 15
#| fig.height: 10
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30))
```

### **Adjusting the rotated x-axis text label**

```{r}
#| fig.width: 15
#| fig.height: 10
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```

## **4. Plotting Interactive Parallel Coordinates Plot: parallelPlot methods**

`parallelPlot` is an R package for creating interactive parallel coordinates plots using `htmlwidgets` and `d3.js`. In this section, you will learn how to use `parallelPlot` to build interactive parallel coordinates plots.

### **The basic plot**

```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))
parallelPlot(wh,
             width = 320,
             height = 250)
```

### **Rotate axis label**

```{r}
parallelPlot(wh,
             rotateTitle = TRUE)
```

### **Changing the colour scheme**

```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)
```

### **Parallel coordinates plot with histogram**

```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)
```
