---
title: "Hands-on Exercise 5A"
author: "ZHANG Xiaoyu"
date: "May 12, 2026"
date-modified: "May 12,2026"
format: html
code-link: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **Creating Ternary Plot with R**

A **ternary diagram** is a specialized triangular graph used to visualize the relationship between three variables. Unlike standard bar or line charts, this plot is shaped like an equilateral triangle where **each side is scaled from 0 to 1**.

In business analytics, we use ternary diagrams to map out compositional data. It has a wide range of applications.For example,in operational cost analysis,we can use ternary diagrams to analyze the efficiency of different business units by plotting **Labor, Materials, and Overhead** costs.

## 1. **Installing and launching R packages**

In this chapter,2 R packages are used : [**ggtern**](http://www.ggtern.com/)and [**Plotly R**](https://plot.ly/r/):

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

## 2. Data Preparation

For the purpose of this hands-on exercise, the [Singapore Residents by Planning AreaSubzone, Age Group, Sex and Type of Dwelling, June 2000-2018](https://www.singstat.gov.sg/find-data/search-by-theme/population/geographic-distribution/latest-data) data will be used.

```{r}
pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv") 
```

Next, use the ***mutate()*** function of **dplyr** package to derive three new measures, namely: young, active, and old.

```{r}
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)
```

## 3. **Plotting Ternary Diagram with R**

This **ternary plot** visualizes the population structure of different planning areas in Singapore for the year 2015. Each black dot on the chart represents a specific neighborhood or district.

By using three axes, we can see the balance between three age groups at once:

-   **ACTIVE:** The top vertex represents the workforce (ages 20–64).

-   **YOUNG:** The bottom-left vertex represents children and youths (ages 0–19).

-   **OLD:** The bottom-right vertex represents seniors (ages 65 and above).

```{r}
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()
```

The current plot is quite basic. We can enhance it by adding **colors** and **icons** to make our insights clearer and more impactful.

```{r}
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()
```

The static plot makes it difficult to identify specific values, which limits our ability to draw meaningful conclusions. To solve this, we need an **interactive ternary plot**.

```{r}
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```
