---
title: "Hands-on Exercise 3A"
author: "ZHANG Xiaoyu"
date: "April 29, 2026"
date-modified: "April 29,2026"
format: html
code-link: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **Programming Interactive Data Visualisation with R**

## 3.1 Package Installation and Data Importing

> Use the following code chunk to check, install, and load the necessary R packages for interactive graphics ([**ggiraph**](https://davidgohel.github.io/ggiraph/), [**plotly**](https://plotly.com/r/), [**DT**](https://rstudio.github.io/DT/)), data science ([**tidyverse**](https://www.tidyverse.org/)), and multi-plot composition ([**patchwork**](https://patchwork.data-imaginist.com/)).

```{r}
pacman::p_load(ggiraph, plotly, 
               patchwork, DT, tidyverse) 
```

This part we still use the same exam data as Hands-on Exercise 1.

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## **3.2 Interactive Data Visualisation - ggiraph methods**

**ggiraph** is a `ggplot2` extension for creating interactive graphics via three main arguments:

-   **Tooltip**: Displays data info on hover.

-   **Onclick**: Executes JavaScript functions on click.

-   **Data_id**: Associates elements with unique IDs for selection.

### **Tooltip effect with *tooltip* aesthetic**

Using **ggiraph** involves a two-step process:

-   **Step 1:** Create a **ggplot** object using interactive geometries (e.g., `geom_point_interactive`).

-   **Step 2:** Use the `girafe()` function to convert that object into an interactive **SVG** graphic.

```{r}
exam_data$MATHS <- as.numeric(exam_data$MATHS)

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = NAME),
    stackgroups = TRUE, 
    binwidth = 1, 
    method = "histodot") +
  scale_y_continuous(NULL, 
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618
)
```

## 3.3 **Interactivity**

The first three lines create a **tooltip** field by combining the **NAME** and **CLASS** data. Line 7 then uses this new field to display tooltips in the plot.

```{r}
exam_data$tooltip <- c(paste0(     
  "Name = ", exam_data$ID,         
  "\n Class = ", exam_data$CLASS)) 

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = exam_data$tooltip), 
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 8,
  height_svg = 8*0.618
)
```

Code chunk below uses [`opts_tooltip()`](https://davidgohel.github.io/ggiraph/reference/opts_tooltip.html) of **ggiraph** to customize tooltip rendering by add css declarations.

```{r}
tooltip_css <- "background-color:white; #<<
font-style:bold; color:black;" #<<

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(tooltip = NAME),                   
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(    #<<
    opts_tooltip(    #<<
      css = tooltip_css)) #<<
)                                        
```

This code customizes tooltips by calculating 90% confidence intervals of the mean and displaying these statistics when hovering over the plot.

```{r}
tooltip <- function(y, ymax, accuracy = .01) {
  mean <- scales::number(y, accuracy = accuracy)
  sem <- scales::number(ymax - y, accuracy = accuracy)
  paste("Mean maths scores:", mean, "+/-", sem)
}

gg_point <- ggplot(data=exam_data, 
                   aes(x = RACE),
) +
  stat_summary(aes(y = MATHS, 
                   tooltip = after_stat(  
                     tooltip(y, ymax))),  
    fun.data = "mean_se", 
    geom = GeomInteractiveCol,  
    fill = "light blue"
  ) +
  stat_summary(aes(y = MATHS),
    fun.data = mean_se,
    geom = "errorbar", width = 0.2, size = 0.2
  )

girafe(ggobj = gg_point,
       width_svg = 8,
       height_svg = 8*0.618)
```

The second interactive feature of ggiraph

```{r}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(           
    aes(data_id = CLASS),             
    stackgroups = TRUE,               
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618                      
)                                        
```

Css codes are used to change the highlighting effect

```{r}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(data_id = CLASS),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(                        
    opts_hover(css = "fill: #202020;"),  
    opts_hover_inv(css = "opacity:0.2;") 
  )                                        
)                                        
```

There are time that we want to combine tooltip and hover effect on the interactive statistical graph as shown in the code chunk below.

```{r}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(tooltip = CLASS, 
        data_id = CLASS),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(                        
    opts_hover(css = "fill: #202020;"),  
    opts_hover_inv(css = "opacity:0.2;") 
  )                                        
)                                        
```

The code chunk below shown an example of `onclick`.

```{r}
exam_data$onclick <- sprintf("window.open(\"%s%s\")",
"https://www.moe.gov.sg/schoolfinder?journey=Primary%20school",
as.character(exam_data$NAME))

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(onclick = onclick),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618)                                                                             
```

To create coordinated multiple views in **ggiraph**:

-   Use **interactive functions** (like `geom_dotplot_interactive`) for each plot.

-   Use the **patchwork** package to combine these plots.

-   Wrap the combined object in the `girafe()` function to enable **linked highlighting** via a shared `data_id`.

```{r}
exam_data$ENGLISH <- as.numeric(exam_data$ENGLISH)

p1 <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(data_id = NAME),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +  
  coord_cartesian(xlim=c(0,100)) + 
  scale_y_continuous(NULL,               
                     breaks = NULL)

p2 <- ggplot(data=exam_data, 
       aes(x = ENGLISH)) +
  geom_dotplot_interactive(              
    aes(data_id = NAME),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") + 
  coord_cartesian(xlim=c(0,100)) + 
  scale_y_continuous(NULL,               
                     breaks = NULL)

girafe(code = print(p1 + p2), 
       width_svg = 6,
       height_svg = 3,
       options = list(
         opts_hover(css = "fill: #202020;"),
         opts_hover_inv(css = "opacity:0.2;")
         )
       ) 
```

## **3.4 Interactive Data Visualisation - plotly methods**

### **Creating an interactive scatter plot: plot_ly() method**

## Interactive Scatter Plot

::: panel-tabset
### The plot

```{r}

plot_ly(
  data = exam_data,
  x = ~MATHS,
  y = ~ENGLISH
)
```

### The code chunk

```{r}
#| eval: false
plot_ly(
  data = exam_data,
  x = ~MATHS,
  y = ~ENGLISH,
)
```
:::

*Next color* argument is mapped to a qualitative visual variable

```{r}
plot_ly(data = exam_data, 
        x = ~ENGLISH, 
        y = ~MATHS, 
        color = ~RACE)
```

Next use *ggplotly():*

```{r}
p <- ggplot(data=exam_data, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
ggplotly(p)
```

The creation of a coordinated linked plot by using plotly involves three steps:

-   [`highlight_key()`](https://www.rdocumentation.org/packages/plotly/versions/4.9.2/topics/highlight_key) of **plotly** package is used as shared data.

-   two scatterplots will be created by using ggplot2 functions.

-   lastly, [*subplot()*](https://plotly.com/r/subplots/) of **plotly** package is used to place them next to each other side-by-side.

```{r}
exam_data$SCIENCE <- as.numeric(exam_data$SCIENCE)

d <- highlight_key(exam_data)
p1 <- ggplot(data=d, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

p2 <- ggplot(data=d, 
            aes(x = MATHS,
                y = SCIENCE)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
subplot(ggplotly(p1),
        ggplotly(p2))
```

-   A wrapper of the JavaScript Library [DataTables](https://datatables.net/)

```{r}
DT::datatable(exam_data, class= "compact")
```

```{r}
d <- highlight_key(exam_data) 
p <- ggplot(d, 
            aes(ENGLISH, 
                MATHS)) + 
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

gg <- highlight(ggplotly(p),        
                "plotly_selected")  

crosstalk::bscols(gg,               
                  DT::datatable(d), 
                  widths = 5)        

```
