---
title: "Hands-on Exercise 7A"
author: "ZHANG Xiaoyu"
date: "June 3, 2026"
date-modified: "June 4,2026"
format: html
code-link: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **Modelling, Visualising and Analysing Network Data with R**

## **1. Overview**

In this exercise, you will learn how to analyse and visualise network data in R. By the end, you will be able to:

-   Create and manage network graph data using **dplyr**, **lubridate**, and **tidygraph**.

-   Build network visualisations with **ggraph**.

-   Calculate and interpret network metrics.

-   Enhance network graphs using network measures.

-   Create interactive network visualisations with **visNetwork**.

## **2. Installing and launching R packages**

```{r}
pacman::p_load(igraph, tidygraph, ggraph, 
               visNetwork, lubridate, clock,
               tidyverse, graphlayouts, 
               concaveman, ggforce)
```

## **3. The Data**

### **3.1 Importing network data from files**

```{r}
GAStech_nodes <- read_csv("data/GAStech_email_node.csv")
GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")
```

### 3.2 **Reviewing the imported data**

```{r}
glimpse(GAStech_edges)
```

### 3.3 **Wrangling time**

```{r}
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate,
                        label = TRUE,
                        abbr = FALSE))
```

### 3.4 **Wrangling attributes**

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source!=target) %>%
  filter(Weight > 1) %>%
  ungroup()
```

## **4. Creating network objects using tidygraph**

### **4.1 Using `tbl_graph()` to build tidygraph data model**

```{r}
GAStech_graph <- tbl_graph(nodes = GAStech_nodes,
                           edges = GAStech_edges_aggregated, 
                           directed = TRUE)
```

### **4.2 Reviewing the output tidygraph’s graph object**

```{r}
GAStech_graph
```

### **4.3 Changing the active object**

```{r}
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```

## **5. Plotting Static Network Graphs with ggraph package**

### **5.1 Plotting a basic network graph**

```{r}
ggraph(GAStech_graph) +
  geom_edge_link() +
  geom_node_point()
```

### **5.2 Changing the default network graph theme**

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

### **5.3 Changing the coloring of the plot**

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes(colour = 'grey50')) +
  geom_node_point(aes(colour = 'grey40'))

g + theme_graph(background = 'grey10',
                text_colour = 'white')
```

### **5.4 Fruchterman and Reingold layout**

```{r}
g <- ggraph(GAStech_graph, 
            layout = "fr") +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()

```

### **5.5 Modifying network nodes**

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes()) +
  geom_node_point(aes(colour = Department, 
                      size = 3))

g + theme_graph()
```

### **5.6 Modifying edges**

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") +
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 3)

g + theme_graph()
```

## **6. Creating facet graphs**

### **6.1 Working with *facet_edges()***

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)

g + facet_edges(~Weekday)
```

### **6.2 Working with *facet_edges()***

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2) +
  theme(legend.position = 'bottom')
  
g + facet_edges(~Weekday)
```

### **6.3 A framed facet graph**

```{r}
set_graph_style() 

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_edges(~Weekday) +
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

### **6.4 Working with *facet_nodes()***

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_nodes(~Department)+
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

## **7. Network Metrics Analysis**

### **7.1 Computing centrality indices**

```{r}
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
            size=betweenness_centrality))
g + theme_graph()
```

### **7.2 Visualising network metrics**

```{r}
g <- GAStech_graph %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department, 
                      size = centrality_betweenness()))
g + theme_graph()
```

### **7.3 Visualising Community**

```{r}
g <- GAStech_graph %>%
  mutate(community = as.factor(
    group_edge_betweenness(
      weights = Weight, 
      directed = TRUE))) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(
    aes(
      width=Weight), 
    alpha=0.2) +
  scale_edge_width(
    range = c(0.1, 5)) +
  geom_node_point(
    aes(colour = community))  

g + theme_graph()
```

```{r}
g <- GAStech_graph %>%
  activate(nodes) %>%
  mutate(community = as.factor(
    group_optimal(weights = Weight)),
         betweenness_measure = centrality_betweenness()) %>%
  ggraph(layout = "fr") +
  geom_mark_hull(
    aes(x, y, 
        group = community, 
        fill = community),  
    alpha = 0.2,  
    expand = unit(0.3, "cm"),  # Expand
    radius = unit(0.3, "cm")  # Smoothness
  ) + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(fill = Department,
                      size = betweenness_measure),
                      color = "black",
                      shape = 21)
  
g + theme_graph()
```

## **8. Building Interactive Network Graph with visNetwork**

### **8.1 Data preparation**

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```

### **8.2 Plotting the first interactive network graph**

```{r}
visNetwork(GAStech_nodes, 
           GAStech_edges_aggregated)
```

### **8.4 Working with visual attributes - Nodes**

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

### **8.5 Working with visual attributes - Edges**

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to", 
           smooth = list(enabled = TRUE, 
                         type = "curvedCW")) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

### **8.6 Interactivity**

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```
