---
title: "Hands-on Exercise 4A"
author: "ZHANG Xiaoyu"
date: "May 07, 2026"
date-modified: "May 07,2026"
format: html
code-link: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **Visualising Distribution**

In this chapter, we will learn two modern methods for visualising data distribution: **ridgeline plots** and **raincloud plots** using **ggplot2** and its extensions.

## **1. Installing and loading the packages**

```{r}
pacman::p_load(ggdist, ggridges, ggthemes,
               colorspace, tidyverse)
```

## **2. Data import**

In this chapter we still use the *Exam_data.csv*

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## **3. Visualising Distribution with Ridgeline Plot**

There are different ways to create ridgeline plots in R. In this section, we use the **ggridges** package.The package provides two main functions:

**（1）geom_ridgeline()**: uses height values directly
**（2）geom_density_ridges()**: estimates data density before plotting

The ridgeline plot below is created using **geom_density_ridges()**.

::: panel-tabset
### The plot

```{r}
exam_data$ENGLISH <- as.numeric(exam_data$ENGLISH)
ggplot(exam_data, 
       aes(x = ENGLISH, 
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
    ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

### The code chunk

```{r}
#| eval: false
exam_data$ENGLISH <- as.numeric(exam_data$ENGLISH)
ggplot(exam_data, 
       aes(x = ENGLISH, 
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
    ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```
:::

### **Varying fill colors along the x axis**

::: panel-tabset
### The plot

```{r}
ggplot(exam_data, 
       aes(x = ENGLISH, 
           y = CLASS,
           fill = stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Temp. [F]",
                       option = "C") +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

### The code chunk

```{r}
#| eval: false
exam_data$ENGLISH <- as.numeric(exam_data$ENGLISH)
ggplot(exam_data, 
       aes(x = ENGLISH, 
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
    ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```
:::

### **Mapping the probabilities directly onto colour**

::: panel-tabset
### The plot

```{r}
ggplot(exam_data,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = 0.5 - abs(0.5-stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient", 
                      calc_ecdf = TRUE) +
  scale_fill_viridis_c(name = "Tail probability",
                       direction = -1) +
  theme_ridges()
```

### The code chunk

```{r}
#| eval: false
ggplot(exam_data,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = 0.5 - abs(0.5-stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient", 
                      calc_ecdf = TRUE) +
  scale_fill_viridis_c(name = "Tail probability",
                       direction = -1) +
  theme_ridges()
```
:::

### **Ridgeline plots with quantile lines**

::: panel-tabset
### The plot

```{r}
ggplot(exam_data,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = 4,
    quantile_lines = TRUE) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()
```

### The code chunk

```{r}
#| eval: false
ggplot(exam_data,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = 4,
    quantile_lines = TRUE) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()
```
:::

Instead of using number to define the quantiles, we can also specify quantiles by cut points such as 2.5% and 97.5% tails to colour the ridgeline plot as shown in the figure below.

::: panel-tabset
### The plot

```{r}
ggplot(exam_data,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = c(0.025, 0.975)
    ) +
  scale_fill_manual(
    name = "Probability",
    values = c("#FF0000A0", "#A0A0A0A0", "#0000FFA0"),
    labels = c("(0, 0.025]", "(0.025, 0.975]", "(0.975, 1]")
  ) +
  theme_ridges()
```

### The code chunk

```{r}
#| eval: false
ggplot(exam_data,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = c(0.025, 0.975)
    ) +
  scale_fill_manual(
    name = "Probability",
    values = c("#FF0000A0", "#A0A0A0A0", "#0000FFA0"),
    labels = c("(0, 0.025]", "(0.025, 0.975]", "(0.975, 1]")
  ) +
  theme_ridges()
```
:::

## **4. Visualising Distribution with Raincloud Plot**

A Raincloud Plot is a way to show data distribution. It shows a half-density curve that looks like a raincloud.

It improves the boxplot because it shows where data is concentrated. It can also show if there are different groups in the data. A boxplot cannot show this clearly.

In this section, we will use **ggdist** and **ggplot2** to make a raincloud plot of English scores by race.

::: panel-tabset
### The plot

```{r}
ggplot(exam_data, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA)
```

### The code chunk

```{r}
#| eval: false
ggplot(exam_data, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA)
```
:::

Next, we will add the second geometry layer using [`geom_boxplot()`](https://r4va.netlify.app/chap09) of ggplot2. This produces a narrow boxplot. We reduce the width and adjust the opacity.

::: panel-tabset
### The plot

```{r}
ggplot(exam_data, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA)
```

### The code chunk

```{r}
#| eval: false
ggplot(exam_data, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA)
```
:::

Lastly, we use **coord_flip()** from **ggplot2**. It flips the plot horizontally. This makes it look like a raincloud.

We also use **theme_economist()** from **ggthemes**. It makes the plot look clean and professional.

::: panel-tabset
### The plot

```{r}
ggplot(exam_data, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 1.5) +
  coord_flip() +
  theme_economist()
```

### The code chunk

```{r}
#| eval: false
ggplot(exam_data, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 1.5) +
  coord_flip() +
  theme_economist()
```
:::
